---
title: "RMarkdown & Summary statistics"
output: html_notebook
---

# Introduction
Jacobo de la Cuesta-Zuluaga. July 2019.

This notebook contains the code and exercises used in the session *Basic statistics*, tutored by Jacobo de la Cuesta-Zuluaga as part of the R for Bioscientists 2019 summer course at the Max-Planck Institute for Developmental Biology.

## Before we begin

There are many good resources on the internet for learning both statistics and R (or both combined!). Don't ever be afraid or ashamed of googling things, this is a skill on its own, and mastering it will help you a lot. Here are some resources that may be of help:

  * https://www.datacamp.com/ : free and paid interactive data analysis and R courses
  * http://www.sthda.com/ : large ammount of tutorials for any kind of analyses and graph generation in R
  * https://www.statmethods.net/index.html : short and clear explanation of R functions
  * https://stats.stackexchange.com/ : forum where people ask and answer all kind of statistics-related questions.
  * https://stackoverflow.com/ : similar to StatsExchange, but for programming-related questions. You will get to love this site, seriously. If you want to do anything in R (or any other programming language), or you keep bumping into an error, just google it; most likely someone has already asked that question in StackOverflow and you will likely get one (or more) answers.
* https://www.r-bloggers.com/ : an aggregator of people bloging various R stuff. You can search for tutorials, news, opinions and other R-related things.
* http://jef.works/R-style-guide/ : Best practices for readable, sharable, and verifiable R code

## Libraries
```{r Load Libraries}
library(skimr)
```

# Summary statistics
This section is adapted from the *Descriptive Statistics and Graphics* of [STHDA](http://www.sthda.com/english/wiki/descriptive-statistics-and-graphics) 

First, let's load the tables we will use and explore their contents.
```{r Load gapminder}
murders = read.delim("/Users/jdelacuesta/Nextcloud/Documents/R_for_bio/2019_Data_Wrangling/2019_Intro/Data/murders.tsv")
```


As you may already know, you can see the top of a table with the function `head()`, and we can see the structure of the dataset with the function `str()`. Let's visualize it to have a glimpse of our data.
```{r Head}
head(murders, n = 10)
str(murders)
```

R contains multiple functions to describe the data. They all work in a very similar fashion: the name of the function describes the statistic you want to obtain, and the parameter it requires is a vector or column of data. Some of the most used are:

* Mean:  `mean()`
* Standard deviation:  `sd()`
* Variance:  `var()`
* Minimum:   `min()`
* Maximum:   `maximum()`
* Median:  median()`
* Range of values (minimum and maximum):   `range()`
* Sample quantiles:  `quantile()`
* Generic function:  `summary()`
* Interquartile range:   `IQR()`

In the present session we will see how these functions work. This will be somewhat repetitive and not too exciting, but it is part of all statistical analyses and you have to learn the name of these functions, there is no other way arround it.

## Shape of the data

One of the first things to know is how many rows and columns our data has. You already got a summary with the `str()` command, but there are ways of getting this information as variables. 

```{r}
# Dimension
dim(murders)

# Number of rows
nrow(murders)

# Number of columns
ncol(murders)
```


## Measures of central tendency: mean, median, mode
```{r Mean Example 1}
# Compute the mean of the population
mean(murders$population)
```

We can verify the result calculating the mean 'manually'
```{r Mean Example 2}
pop_sum = sum(murders$population)
n_pop = length(murders$population)
pop_sum/n_pop
```

```{r Mean Exercise}
# Compute the mean number of murders

```

The median is calculated in a similar way
```{r Median Example}
# Compute the median population
median(murders$population)
```

## Measures of variablity

To determine the spread of the data, you can calculate the range, that is, the minimum and the maximum
```{r Range Example}
range(murders$population)
```

If you want just one of those values, you can use the `min()` and `max()` functions. 

*exercise:* Obtain the minimum and the maximum number of murders, create a vector with these values, and compare it with the result you get from the `range()` function
```{r Range Exercise}

```

*exercise:*  You may recall that the standard deviation equals the square root of the variance. Using the functions to compute the variance and the standard deviation of a sample (that you can find in the list above) obtain the standard deviation from the variance, and the variance from the standard deviation.
```{r SD Exercise}

```

If you are dealing with data that don't distribute normally, it is better to use metrics that better accoun for the skewness. In this case, it is better to use the median and the interquartile range. 

## Dealing with non-normal data

*exercise:* Obtain the median and interquartile range of both the state population and the number of murders. Compare these values with the mean and standard deviation.
```{r Median Exercise}
# Compute the median and iqr of number of murders

# Compute the median and iqr the state population

# Compute the mean and sd of number of murders

# Compute the mean and sd the state population
```

## Other statistics

You may also want to cut the distribution of your data in other points. If we were interested in determininig the value that divide the states by population in the top 10% and bottom 90%, we can use the `quantile()` function.

```{r Example Quantile 1}
quantile(murders$population, probs = 0.9)
```
It is also possible to calculate multiple values simultaneously, for example, if we wanted to obtain the top and bottom 10%, we pass a vector of probabilities to the function

```{r Example Quantile 2}
quantile(murders$population, probs = c(0.1, 0.9))
```

*exercise:* Using the results of the last exercise, manually calculate the median and interquartile range of the number of murders and population using `quartile()`.
```{r Exercise Quantile 3}

```


## Summarizing data

While this is not very difficult, it can also be handy to obtain several of these statistics using a single function. For this we have `summary()`, which works just like all we just saw above, and which returns a vector we can save or index.

```{r Example Summary}
# Summary of the number of murders and population 
summary(murders$population)

summary(murders$total)

```

## Aggregating data

Now that we have covered the basics, we can do something a little more advanced. If you check again the structure of the murders data frame, you can see that there is a categorical variable (also called factor) which groups states by region. We may ask how do the values of murders and population vary per region. You could do it the long way, by subseting the dataframe by region and calculating every statistic on each dataframe, or you could use the `aggregate()` function, which does exactly that for you.

Aggregate uses the formula notation of R, which works as follows: `y ~ x` and is read `y` *by* `x`. In the following example, we will calculate the mean population *by* region. You can also combine two categorical variables in your formula as follows: `y ~ x + z`

```{r Example Aggregate 1}
# Aggregate
aggregate(total ~ region, data = murders, FUN = mean)
```
Note the other two parameters needed by the function: `data` tells the function where to look for the variables we are including in the formula, and `FUN` tells which function to calculate.

If were to make a table for a paper with these information, we would also need the standard deviation. It would be just a matter of changing the parameter in `FUN`
```{r Example Aggregate 2}
# Aggregate
aggregate(total ~ region, data = murders, FUN = sd)
```

*exercise:* Calculate the range of murders by region
```{r Exercise Aggregate }
# Aggregate

```

## Tabulate data

So far we have been doing summaries of continuous data, but we can also summarize categorical varaibles in R with a couple of handy functions.

For the following examples, we will use `HairEyeColor`, a dataset included in R by default, so there is no need to load any file. This data set contains information of the frequency of  hair  and eye color by sex in a group of 592 students.

First, we need to convert the data into a long data frame as follows:
```{r Convert HairEyeColor}
hair_eye_col = as.data.frame(HairEyeColor)
HE_color = hair_eye_col[rep(row.names(hair_eye_col), hair_eye_col$Freq), 1:3]

head(HE_color)
```

First, we can obtain tables of frecuencies (or counts) of each of the variables using the function `table()`
```{r Example Table 1}
# Table
table(HE_color$Hair)
# Convert to data frame
as.data.frame(table(HE_color$Hair))
```

*exercise:* determine the number of students by sex and  by eye color (separately)
```{r exercise Table 1}
# Table sex and age

```

We can also create contingency tables with the same function by passing two variables. For example, if we want to know the number of students by sex and eye color

```{r Example Table 2}
table(HE_color$Sex, HE_color$Eye)
```

To create tables according to multiple categorical variables, you can use `ftable()`, which takes as input a formula of the form `Cat1 + Cat2 ~ Cont`
```{r Example Ftable}
ftable(Sex + Hair ~ Eye, data = HE_color)
```

You can use functions to get information from table objects. For example, if you are interested in the margin sums of a table, you can use the `margin.table()` function
```{r Example Margintable}
# Table of Eye and Hair color
HE_table =  table(HE_color$Hair, HE_color$Eye)
HE_table
# Margin 1 means by rows  
margin.table(HE_table, margin = 1) 

# Margin 2 means by columns
margin.table(HE_table, margin = 2)
```

Finally, you can obtain a table of proportions using the `prop.table()` function. You should also specify whether 
```{r Example Proptable}
prop.table(HE_table)
```

*exercise:* Determine whether `prop.table()` uses rows or columns totals to calculate the proportions. Now compute a proportion table in the oposite sense (i.e. if defaults if rows use columns, or if the default is columns, use rows); verify it.

```{r exercise Proptable}

```

## Extra: summaries with Skimr

There is a very convinient R package called `skimr`, which can create summary statistics of a table. We'll se its main function. Remember first to load the package at the top of the notebook.

To generate summaries of a complete data frame, you use the function `skim()`

```{r Example Skim}
skim(murders)
```

*exercise:* Explore the output table. How are the variables separated? What values does the function provide? For each of the values, remember the function of base R used to calculate it.

```{r exercise Skimr 1}

```

This looks pretty, but may be difficult to use in downstream analyses. To obtain a more R friendly table, the package also provides a funtion called `skim_to_wide()`, which returns a regular data frame.

*exercise:* create a wide summary of the `murders` data frame. How is it different to the original skim summary?

```{r exercise Skimr 2}

```

