---
title: "RMarkdown & Summary statistics"
output: html_notebook
---

# Introduction
Jacobo de la Cuesta-Zuluaga. July 2019.

This notebook contains the code and exercises used in the session *Basic statistics*, tutored by Jacobo de la Cuesta-Zuluaga as part of the R for Bioscientists 2019 summer course at the Max-Planck Institute for Developmental Biology.

## Before we begin

There are many good resources on the internet for learning both statistics and R (or both combined!). Don't ever be afraid or ashamed of googling things, this is a skill on its own, and mastering it will help you a lot. Here are some resources that may be of help:

  * https://www.datacamp.com/ : free and paid interactive data analysis and R courses
  * http://www.sthda.com/ : large ammount of tutorials for any kind of analyses and graph generation in R
  * https://www.statmethods.net/index.html : short and clear explanation of R functions
  * https://stats.stackexchange.com/ : forum where people ask and answer all kind of statistics-related questions.
  * https://stackoverflow.com/ : similar to StatsExchange, but for programming-related questions. You will get to love this site, seriously. If you want to do anything in R (or any other programming language), or you keep bumping into an error, just google it; most likely someone has already asked that question in StackOverflow and you will likely get one (or more) answers.
  * https://www.r-bloggers.com/ : an aggregator of people bloging various R stuff. You can search for tutorials, news, opinions and other R-related things.
* http://jef.works/R-style-guide/ : Best practices for readable, sharable, and verifiable R code

## Libraries
```{r Load Libraries}

```

# Markdown
```{r}

```

# Summary statistics
This section is adapted from the *Descriptive Statistics and Graphics* of STHDA, available from http://www.sthda.com/english/wiki/descriptive-statistics-and-graphics

First, let's load the tables we will use and explore their contents.
```{r Load gapminder}
murders = read.delim("/Users/jdelacuesta/Nextcloud/Documents/R_for_bio/2019_Data_Wrangling/Data/murders.tsv")
str(murders)
```

```{r Head}
head(murders, n = 10)
```

There are multiple R functions to describe the data:

* Mean:  `mean()`
* Standard deviation:  `sd()`
* Variance:  `var()`
* Minimum:   `min()`
* Maximum:   `maximum()`
* Median:  median()`
* Range of values (minimum and maximum):   `range()`
* Sample quantiles:  `quantile()`
* Generic function:  `summary()`
* Interquartile range:   `IQR()`

## Measures of central tendency: mean, median, mode
```{r Mean Example 1}
# Compute the mean of the population
mean(murders$population)
```

We can verify the result calculating the mean 'manually'
```{r Mean Example 2}
pop_sum = sum(murders$population)
n_pop = length(murders$population)
pop_sum/n_pop
```

```{r Mean Excercise}
# Compute the mean number of murders

```

The median is calculated in a similar way
```{r Median Example}
# Compute the median population
median(murders$population)
```

```{r Median Excercise}
# Compute the median number of murders

```

## Measures of variablity

To determine the spread of the data, you can calculate the range, that is, the minimum and the maximum
```{r Range Example}
range(murders$population)
```

If you want just one of those values, you can use the `min()` and `max()` functions. 

*Excercise:* Obtain the minimum and the maximum number of murders, create a vector with these values, and compare it with the result you get from the `range()` function
```{r Range Excercise}

```
