---
title: "Data graphics with ggplot2"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Why data graphics? 

## What kind of graph to use?

## How to plot it?

 * Base R `graphics`
 * `lattice` and `grid`
 * `ggplot2`

## General rules for data graphics

 1. Display data values using position or length
 2. Use horizontal lengths in preference to vertical lengths
 3. Watch your data-ink ratio
 4. Think very carefully before using color to represent data values

### "Do nots"

 5. Do *not* use areas to represent data values
 6. *Please* do not use angles or slopes to represent data values
 7. *Please, please* do not use volumes to represent data values

Source: [Paul Murrell, "Comment"](https://www.stat.auckland.ac.nz/~paul/MurrellCommentGelmanUnwin.pdf)

## Tables vs Graphics

## Elements of a plot

| Element                    | Function name prefix |
|----------------------------|----------------------|
| Data                       | `ggplot(data=...)`   |
| Aesthetic mapping          | `ggplot(aes(...))`   |
| Geometric object           | `geom_...`           |
| Statistical transformation | `stat_...`           |
| Scaling                    | `scale_...`          |
| Coordinate system          | `coord_...`          |
| Facetting                  | `facet_...`          |

## Aesthetic mappings

```{r echo=TRUE}
data(mtcars)
```

## Are these colors "evenly spaced"?

```{r}
palette(rainbow(6, # presented by Tom Clancy
                start=0)) 
barplot(rep(1,times=6),col=1:6, axes=FALSE)
```

## Be careful with colors

 * Perception is influenced by both hue and intensity
 * Use color-blind safe palettes where possible
 * Avoid using color for quantitative scales

## Where to put the cutoff? 

```{r echo=FALSE}
data(diamonds)
(ggplot(diamonds, aes(x=depth, y=price)) 
 + geom_density_2d() 
 + stat_density_2d(geom="raster", aes(fill=stat(density)), contour=FALSE)
 + scale_fill_gradientn(colors=rainbow(6,start=0)) # Destroy your retinas
 + scale_y_log10() # Make the spread of values easier to see
 )
```

## What about now?

```{r echo=FALSE}
(ggplot(diamonds, aes(x=depth, y=price)) 
 + geom_density_2d() 
 + stat_density_2d(geom="raster", aes(fill=stat(density)), contour=FALSE)
 + scale_y_log10() # Make the spread of values easier to see
 )
```


## Building a plot

 1. Get data object
 2. Check data structure
 3. Decide what to plot, and what aesthetic mapping to use
 4. Choose the geometric object(s)
 5. Adjust the scales if necessary
 6. Facet by factors if necessary
 7. Add labels to the plot

## Get data object. Check data structure
 
```{r echo=TRUE}
data(mtcars)
str(mtcars)
```

## Decide what to plot, and what aesthetic mapping to use

```{r echo=TRUE}
mtcars.plot <- ggplot(mtcars, aes(x=hp,y=mpg,col=cyl))
```

## Choose the geometric object(s)

```{r echo=TRUE}
mtcars.plot + geom_point()
```

## Adjust the scales if necessary

```{r echo=TRUE}
mtcars.plot + geom_point() + scale_y_log10()
```

## Facet by factors if necessary
 
```{r echo=TRUE}
mtcars.plot + geom_point() + scale_y_log10() + facet_wrap(~cyl)
```

"Wait, didn't you say facet by **factors**?"

## Add labels to the plot

```{r echo=TRUE}
(mtcars.plot + geom_point() 
 + labs(title="motorcars",x="horsepower",
        y="miles per gallon",col="cylinders"))
```

## Now it's your turn to try... 

 * Import data files in `R4eBio/2019_Intro/Data/` 
 * Scatter plot murders vs. population (`murders.tsv`)
   * Bored? Color by region, facet by region, label by state abbreviation (hint: `geom_text()`)
   * Still bored? Add a trend line (hint: `geom_smooth()`), choose a linear model
 * Histogram, violinplot, or boxplot of heights (`heights.tsv`)
   * Bored? Color by sex, facet by sex (in one row, then one column)
 * Line plot infant mortality by country per year (`gapminder.tsv`) (hint: `geom_line()` and `group` aesthetic)
   * Bored? Make lines transparent, color by continent, overlay with trend lines (one per continent), facet by continent

## Learn more

 * [ggplot2 Function Reference](https://ggplot2.tidyverse.org/reference/index.html)
 * `Reference_notebook_ggplot2.Rmd` in the `R4eBio/2019_Intro/Notebooks/` folder - worked examples and links to more resources