---
title: "Homework - Data wrangling and tidying with the tidyverse"
output: html_notebook
---

# Introduction
Jacobo de la Cuesta-Zuluaga. July 2019.

The present notebook contains excercises about basic data wrangling using the tidyverse. It is base on the material of Lisa DeBruine and Dale Barr's [MSC Data Skills course](https://psyteachr.github.io/msc-data-skills/)

## Load data and libraries

For the excercises you will use a data table we haven't used before, the UK baby names data set. It is located in the same folder as the ones we used in class.

Load the file and call it `ukbabynames`
```{r}
ukbabynames = NULL
```

Load the libraries
```{r}

```

## Questions

### Question 1

What is the number of entries in the data set? There are several ways of doing this, but the answer in this case should be a single number saved in a variable

```{r Q1}
n_entries = NULL
```

### Question 2

Create a table with the names and the ranks of the babies born in 1999

```{r Q2}
name_ranks = NULL
```

### Question 3

What is the range of birth years contained in the dataset? Use `summarise` to make a table with two columns: `minyear` and `maxyear`.

```{r Q3}
birth_range = NULL
```

### Question 4

Make a table of only the data from babies named Hermione. When was it most popular?

```{r Q4}
hermiones = NULL
```

### Question 5

Sort the dataset by sex and then by year (descending) and then by rank (descending).

```{r Q5}
sorted_babies = NULL
```

### Question 6

Create a new column, `decade`, that contains the decade of birth (1990, 2000, 2010). This one is tricky, be creative! Hint: see `?floor` or `?round`

```{r Q6}
ukb_decade = NULL
```

### Question 7

Make a table of only the data from male babies named Courtney that were born between 1998 and 2001 (inclusive).

```{r Q7}
courtney = NULL
```


### Question 8

How many distinct names are represented in the dataset? Make sure `distinct_names` is an integer, not a data table. Hint: see `?n_distinct` and note it takes a vector of values as input.

```{r Q8}
distinct_names = NULL
```

### Question 9

Make a table of only the data from the female babies named Frankie that were born before 1999 or after 2010.

```{r Q9}
frankie = NULL
```

### Question 10

How many total babies in the dataset were named 'Emily'? Make sure `emily` is an integer, not a data table.

```{r Q10}
emily = NULL
```

### Question 11

How many distinct names are there for each sex? Hint: this need a function you just learned about.

```{r Q11}
names_per_sex = NULL
```

### Question 12

What is the most popular name in the dataset? A simple call to `arrange()` won't be enough!

```{r Q12}
most_popular = NULL
```

### Question 13

How many babies were born each year for each sex?

```{r Q13}
babies_per_year = NULL
```
