---
title: "Basic plotting with ggplot2 - Homework"
output: html_notebook
---

```{r}
library(tidyverse)
```

# Homework exercises

## Data visualization concepts

"Food for thought" questions - no strictly right or wrong answers, just to get 
you to start thinking about how data visualization works.

1. Explain a histogram in terms of the graphical elements listed above.

2. Explain Florence Nightingale's 
 [coxcomb diagram](http://www.florence-nightingale-avenging-angel.co.uk/Coxcomb.tif) 
 in terms of the graphical grammar. 
 [Further reading](http://www.florence-nightingale-avenging-angel.co.uk/?p=462)

3. How is scaling different from a coordinate system?

## Scatterplots 

Load the `gapminder.tsv` data from Jacobo for the following exercises.

```{r}
gapminder <- read.delim("../Data/gapminder.tsv")
```

1. What is the relationship between fertility and population? 
Plot a scatterplot of fertility vs population for all countries for 1970. 
Save the plot to a named ggplot object, e.g. `fertility_population_1970`. 

```{r}

```

2. In the plot above, also color the points by continent.

```{r}

```

3. In the plot above, also scale the point sizes by gdp. 

Note the error message, which did not appear before. 
What does it mean? 
How might you preprocess the data to deal with this? Should you?

```{r}

```

> The error message only appeared when the variable `gdp` was added to the plot. 
> This probably means that the column `gdp` in the data.frame has some missing 
> values, which were omitted from the plot. 

> One way to deal with this is to filter the table before plotting. However, 
> when exploring the data for the first time, it is usually worth checking which 
> values are missing. This may indicate problems with the upstream data 
> processing, or it may be the relevant to the analysis or interpretation of 
> your data and its reliability.

4. The points are not evenly spaced in the plot, but skewed to one side. 

How might you improve legibility? (Hint: This is a matter of scaling...)

```{r}

```


5. Export the plot as a PDF file, 10 by 5 cm in size.

```{r}

```

## Line plots

We continue to use the `gapminder` data for this exercise (see above).

1. How has life expectancy changed over time? Plot the life expectancy (vertical
axis) vs. year (horizontal) with `geom_line`, drawing one line per country.
(Reminder: Use the `group` aesthetic.)

```{r}

```

2. The above is not very legible or informative. Color and facet the lines by 
continent to help the reader see some trends. (Hint: Use facet_wrap)

Does it make more sense to have the facets in a single row or a single column?

```{r}

```

3. *Challenge* - Superimpose a summary trend line for each continent. 
(Hint 1: Look up the ggplot2 reference to find another geom that draws trend 
lines.)
(Hint 2: Lighten the individual lines with the `alpha` parameter in the 
`geom_line()` function.)

```{r}

```

## Barplots

We use the Titanic dataset for this exercise.

```{r}
# Run this first to reformat the Titanic dataset 
# Load data
data(Titanic)
# Convert to data.frame
titanic.df <- as.data.frame(Titanic)
# Convert to long format
titanic.long <- titanic.df[rep(row.names(titanic.df),titanic.df$Freq),1:4]
```

1. Produce a barplot of counts of `Survived` from the Titanic dataset, facetted 
by both `Sex` and `Class`

```{r}

```

2. Produce the same barplot as above, but with horizontal bars. (Hint: This is
not done with `aes(y=...)`, but with a coordinate transformation function.)

```{r}

```

## Histograms, violin plots, and boxplots

We continue to use the `gapminder` data for this section's exercises.

1. Conceptual question: What are some advantages of violinplots vs. boxplots? 
What are some disadvantages? 
(Hint: consider the shape of a violin.)

## Additional exercises: Base R graphics

Although ggplot2 is a powerful and versatile package, it is still useful to 
know how to plot in base R graphics.

Exercises 1 and 2 use the `iris` dataset.

```{r}
data(iris)
```

1. Plot `Sepal.Length` vs. `Petal.Width`, and add a title and axis labels.

```{r}

```

2. Add title and axis labels to the plot. (Hint: see help page for `plot` 
function)

```{r}

```

Exercises 3, 4, and 5 use the following simulated data.

```{r}
df.random <- data.frame(x=1:100,y=rnorm(100))
```

3. Plot `df.random$y` vs. `df.random$x` as a line instead of points. 
(Hint: `type` parameter)

```{r}

```

4. Annotate the above plot with a horizontal line at y = 0.5. 
(Hint: look up the `abline` function)

```{r}

```

5. Draw a histogram of `df.random$y`. (Hint: look up the `hist` function.) 

```{r}

```

6. Save the histogram plot as a PDF file.

```{r}

```

